Tools for Identifying Partisan Gerrymandering

### 2019-06-18 Latex/_maintext.tex
Update _maintext.tex

    To facilitate comparisons to the two most politically important remedy plans offered to the Court, the joint plan proposed by the Republicans leaders of the two legislative chambers, and the Governor’s proposed plan as a Democratic alternative, we look at a number of features in each of the four plans as a whole. We do not, however, discuss district-specific criteria for identifying gerrymandering in these maps, since, although some district-specific evidence was presented about the 2011 map in the trial testimony by an expert who was highly knowledgeable about communities of interest in the state, the Pennsylvania Supreme Court opinion itself offers a \underline{statewide} standard for a finding of unconstitutional partisan gerrymandering. 

@@

        \footnote{As noted earlier, given the language in the U.S. Supreme Court’s remand of the \textit{Gill} case, it appears that any federal standard for \underline{partisan gerrymandering} would need to be district specific.}

@@

    Rebuttal to any claims that the partisan asymmetries observed in the consistent outcomes of the 2011 map can be attributed primarily to differential concentration of party voting strengths, or to chance, is found in the expert witness testimony in the case (see esp. the simulations of potential good government maps done by Professor Chen, discussed in detail in the Pennsylvania Supreme Court opinion in the case, and on which the Court placed great reliance). With respect to the durability of the effects of the 2011 map, we have already pointed out that this plan yielded unchanged 13R-5D results for three elections in a row. To help explain this over time consistency we note that the 2011 Map has primarily safe seats (13 of 18 won with greater than a 10 percentage point margin). But, indicative of the pro-Republican gerrymander achieved with the 2011 map, those safe seats are distributed in a highly asymmetric fashion, with nine essentially safe seats for Republicans and only four for Democrats.

@@

    There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) \textit{declination} \citep{Warrington2018} (details below). Each of these measures can be directly calculated for the 2011 map based on district election outcomes in 2012, 2014, and 2016.

@@

    \textit{Partisan Bias} is then calculated from the intercept of this regression using an exponential transformation (for details see \citet{Grofman1983}). Partisan bias is calculated from the Republican perspective, and \underline{negative} values indicates bias against Republicans.

@@

    In order to assess expected partisan consequences for plans that have not yet been implemented we need to develop a predictive equation, and one such way is with data on partisan outcomes for state-wide elections projected into the new districts. For example, Professor Jowei Chen in his testimony in the Pennsylvania case used a composite of the six state-wide elections between 2008 and 2010 to estimate the political consequences of alternative maps. After extensive investigation of alternative prediction models and their accuracy in reproducing the partisan outcomes of the 2012, 2014, and 2016 congressional elections at the district level (data omitted for space reasons), we have opted for a model that takes the average of five state-wide 2016 elections that reflect a balance of Republican and Democratic success: President, U.S. Senate, PA Attorney General, PA Treasurer, and PA Auditor. Republicans won the statewide vote in two of them (President, U.S. Senate), while the Democrats won the statewide vote in three of them (Auditor, Treasurer, and Attorney General). The average vote for the Republicans across these five elections was 49.1\%.

@@

    We are not claiming to predict vote shares in advance of an election. We also recognize that the composite of state-wide elections is intended to provide a ``normal vote'' baseline that must be ``corrected'' with incumbency effects if we wish to actually predict outcomes.

@@

    We wish to do a further robustness check on the plausibility of using our five-election composite to calculate the various metrics designed to measure potential gerrymandering. One way to do this is to see how well the composite (calculated at a hypothetical 50\% vote share) mimics the results of the values for the four metrics for the 2011 map that we previously obtained using the actual congressional election results for that map. What we are doing is comparing a hypothetical ``normal vote'' \citep{Converse1966} assumption (based on a statewide vote in which we are not taking incumbency effects into account) with the average of what we actually get in three different congressional elections. The congressional results for 2012-2016 and an average of the three, reported in Table \ref{tab:congsum}, are 0.11 for partisan bias, 0.16 for the efficiency gap, 0.06 for the value of the mean minus median gap, and 0.40 for \textit{declination}. The comparable results from Table \ref{tab:gerry} (see below) are 0.09, 0.24, 0.05, and 0.39. It is apparent how close these two sets of values are to one other even though one is based on actual elections with a mean Republican (two-party) vote share close to 49.1\% while the other is based on projected election results set to yield a 50-50 two-party vote distribution. If instead we calculate these four metrics using the composite five election statewide vote in these elections over the period 2012-2016, a Democratic vote share of 49.1\% we get virtually identical values: 0.09, 0.24, 0.05, 0.36.

@@

    The findings from Table \ref{tab:gerry} are very clear; regardless of which metric we examine, the 2011 map is the most gerrymandered.
    Not surprisingly, the nature of the bias is in a pro-Republican direction. Given the cumulative weight of all the evidence, the 2011 congressional map was clearly an egregious pro-Republican gerrymander. All four measures of potential gerrymandering are statistically significant for the 2011 map. In contrast, as we might expect, the court-ordered plan, prepared by a respected expert whose instructions were to satisfy good government criteria, was by a considerable margin the closest to a perfectly neutral plan in partisan terms according to all four measures. Indeed, none of the four measures of potential gerrymandering is statistically significant for the court-ordered map. And when we turn to the Republican Joint Legislative plan we find that all four measures are again statistically significant.  

@@

    In 2018, even under a court-drawn plan, in a year when Democrats had won a clear majority of the aggregate two-party congressional vote, Democrats only succeeded in winning exactly 50\% of the seats. Trying to understand why evan plans not drawn by Republicans had some level of pro-Republican bias requires us to look in more detail at the partisan electoral geography of Pennsylvania. In Pennsylvania, Philadelphia is overwhelmingly Democratic in voting. In particular, if you draw two congressional districts entirely within Philadelphia County, one of them is very likely to give Democratic candidates around 90\% of the vote, and for sure, as long as both are wholly within the County, the average vote in the two will be around 80\% Democratic no matter how you draw the two districts. There are no other large, concentrated pockets of equivalently overwhelmingly Republican voting strength. Thus, more Democratic votes will \textit{naturally} be ``wasted'' in Philadelphia than Republican votes will be ``wasted'' elsewhere in the state. Additionally, the city of Philadelphia is landlocked, and its position on the state's border reduces the ability of line drawers who favor Democrats to distribute these voters into multiple districts so as to minimize the extent to which Democratic votes in the county are wasted. Additional Democratic wasted votes come in heavily Democratic Allegheny County, the county in which the city of Pittsburgh is located. However, saying that urban concentration of Democrats means that it is harder for Democrats to turn their votes into seats than is the case for Republicans does not mean that we cannot distinguish the effects of such \textit{natural} gerrymanders from \textit{intentional} political gerrymandering, since the latter generates perverse consequences for Democratic voters to a far greater extent than the likely one district or so penalty created by the nature of Pennsylvania’s present electoral geography. 

>>>>

    To facilitate comparisons of the 2011 enacted and 2018 court plans to the two most politically important remedy plans offered to the Court, the joint plan proposed by the Republicans leaders of the two legislative chambers, and the Governor’s proposed plan as a Democratic alternative, we look at a number of features in each of the four plans as a whole. We do not, however, discuss district-specific criteria for identifying gerrymandering in these maps, since, although some district-specific evidence was presented about the 2011 map in the trial testimony by an expert who was highly knowledgeable about communities of interest in the state, the Pennsylvania Supreme Court opinion itself offers a \underline{statewide standard} for a finding of unconstitutional partisan gerrymandering. 

@@

        \footnote{As noted earlier, given the language in the U.S. Supreme Court’s remand of the \textit{Gill} case, it appears that any \underline{federal} standard for \underline{partisan gerrymandering} would need to be district specific.}

@@

    Rebuttal to any claims that the partisan asymmetries observed in the consistent outcomes of the 2011 map can be attributed primarily to differential concentration of party voting strengths, or to chance, is found in the expert witness testimony in the case (see esp. the simulations of potential good government maps done by Professor Chen, discussed in detail in the Pennsylvania Supreme Court opinion in the case, and on which the Court placed great reliance). With respect to the durability of the effects of the 2011 map, this plan yielded unchanged 13R-5D results in all three elections where the map was used. To help explain this over time consistency we note that the 2011 Map has primarily safe seats (13 of 18 won with greater than a 10 percentage point margin). But, indicative of the pro-Republican gerrymander achieved with the 2011 map, those safe seats are distributed in a highly asymmetric fashion, with nine essentially safe seats for Republicans and only four for Democrats.

@@

    There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) \textit{declination} \citep{Warrington2018}. Each of these measures can be directly calculated for the 2011 map based on district election outcomes in 2012, 2014, and 2016.

@@

    \textit{Partisan Bias} is then calculated from the intercept of this regression using an exponential transformation (for details see \citet{Grofman1983}).

@@

    In order to assess expected partisan consequences for plans that have not yet been implemented we need to develop a predictive equation, and one such way is with data on partisan outcomes for state-wide elections projected into the new districts. For example, Professor Jowei Chen in his testimony in the Pennsylvania case used a composite of the six state-wide elections between 2008 and 2010 to estimate the political consequences of alternative maps. After extensive investigation of alternative prediction models and their accuracy in reproducing the partisan outcomes of the 2012, 2014, and 2016 congressional elections at the district level (data omitted for space reasons), we have opted for a model that takes the average of five state-wide 2016 elections that reflect a balance of Republican and Democratic success: President, U.S. Senate, PA Attorney General, PA Treasurer, and PA Auditor.
    Republicans won the statewide vote in two of them (President, U.S. Senate), while the Democrats won the statewide vote in three of them (Auditor, Treasurer, and Attorney General). The average vote for the Republicans across these five elections was 49.1\%.

@@

    We are not claiming to predict vote shares in advance of an election. We also recognize that the composite of state-wide elections is intended to provide a ``normal vote'' baseline that must be ``corrected'' with incumbency effects if we do wish to actually predict outcomes.

@@

    We wish to do a further robustness check on the plausibility of using our five-election composite to calculate the various metrics designed to measure potential gerrymandering. One way to do this is to see how well the composite (calculated at a hypothetical 50\% vote share) mimics the results of the values for the four metrics for the 2011 map that we previously obtained using the actual congressional election results for that map. What we are doing is comparing a hypothetical ``normal vote'' \citep{Converse1966} assumption (based on a statewide vote in which we are not taking incumbency effects into account) with the average of what we actually get in three different congressional elections. The congressional results for 2012-2016 and an average of the three, reported in Table \ref{tab:congsum}, are 0.11 for partisan bias, 0.16 for the efficiency gap, 0.06 for the value of the mean minus median gap, and 0.40 for \textit{declination}. The comparable results from Table \ref{tab:gerry} for the 2011 enacted map at 50\% vote-share are 0.09, 0.24, 0.05, and 0.39.

@@

    The findings from Table \ref{tab:gerry} are very clear; regardless of which metric we examine, the 2011 map is the most gerrymandered. Not surprisingly, the nature of the bias is in a pro-Republican direction. Given the cumulative weight of all the evidence, the 2011 congressional map was clearly an egregious pro-Republican gerrymander. All four measures of potential gerrymandering are statistically significant for the 2011 map. In contrast, as we might expect, the court-ordered plan, prepared by a respected expert whose instructions were to satisfy good government criteria, was by a considerable margin the closest to a perfectly neutral plan in partisan terms according to all four measures. Indeed, none of the four measures of potential gerrymandering is statistically significant for the court-ordered map. And when we turn to the Republican Joint Legislative plan we find that all four measures are again statistically significant.  

@@

    In 2018, even under a court-drawn plan, in a year when Democrats had won a clear majority of the aggregate two-party congressional vote, Democrats only succeeded in winning exactly 50\% of the seats. Trying to understand why evn plans not drawn by Republicans had some level of pro-Republican bias requires us to look in more detail at the partisan electoral geography of Pennsylvania. In Pennsylvania, Philadelphia is overwhelmingly Democratic in voting. In particular, if you draw two congressional districts entirely within Philadelphia County, one of them is very likely to give Democratic candidates around 90\% of the vote, and for sure, as long as both are wholly within the County, the average vote in the two will be around 80\% Democratic no matter how you draw the two districts. There are no other large, concentrated pockets of equivalently overwhelmingly Republican voting strength. Thus, more Democratic votes will \textit{naturally} be ``wasted'' in Philadelphia than Republican votes will be ``wasted'' elsewhere in the state. Additionally, the city of Philadelphia is landlocked, and its position on the state's border reduces the ability of line drawers who favor Democrats to distribute these voters into multiple districts so as to minimize the extent to which Democratic votes in the county are wasted. Additional Democratic wasted votes come in heavily Democratic Allegheny County, the county in which the city of Pittsburgh is located. However, saying that urban concentration of Democrats means that it is harder for Democrats to turn their votes into seats than is the case for Republicans does not mean that we cannot distinguish the effects of such \textit{natural} gerrymanders from \textit{intentional} political gerrymandering, since the latter generates perverse consequences for Democratic voters to a far greater extent than the likely one district or so penalty created by the nature of Pennsylvania’s present electoral geography. 
